Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-12-03 Rversion/reviewR.tex
Propagated changes to R version

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Measuring contrast processing in the visual system using SSVEP},
  pdfauthor={Alex R. Wade; Daniel H. Baker},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Measuring contrast processing in the visual system using SSVEP}
\author{Alex R. Wade \and Daniel H. Baker}
\date{}

@@

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for over a century. One of the most productive
experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high SNR, and they allow us to analyze interactions between
stimulus components using a convenient mathematical framework. Here we
describe how SSVEPs have been used to study visual contrast over the
past 70 years (Dawson, 1954). Because our thinking about SSVEPs is
well-described by simple mathematical models, we embed code that
illustrates key steps in the modelling and analysis. This paper can
therefore be used both as a review of the use of SSVEP in measuring
human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as `contrast', sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity (Campbell \& Green, 1965),
defined as the inverse of the lowest contrast that can be reliably
detected. But neural responses can also be measured more directly using
techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG
method known as the steady state visually evoked potential (SSVEP)
technique has contributed to our understanding of human contrast
processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by
visual stimuli flickering at a constant frequency (Regan, 1966). For
contrast-defined stimuli, such as sine-wave gratings, it is strongest at
the occipital pole, adjacent to the early visual areas that generate the
signal, although careful analysis of individual VEPs reveals multiple
generators throughout visual cortex (Di Russo \emph{et al.}, 2005,
2007). The flickering stimulus entrains neural population responses at
multiples of the stimulus frequency, so continuous EEG data are
typically analysed by taking the Fourier transform, and estimating the
amplitude at these frequencies. Two common variants involve sinusoidal
on-off flicker, where the stimulus alternates between a blank background
and the peak contrast, and sinusoidal counterphase flicker, where the
stimulus alternates in phase (i.e the black regions become white and the
white regions become black). On-off flicker can drive independent
populations of on- and off-cells in the retina once per cycle and can
therefore produce a response at the fundamental flicker frequency, known
as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase
flicker contains two transients per cycle and therefore does not produce
a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on.
Because square-waves are spectrally broad-band, square wave flicker
tends to produce more complex spectral harmonics than sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). SSVEP signals can also be elicited by periodic changes of
stimulus properties other than achromatic and chromatic contrast, such
as motion, stereo depth, and facial identity or expression (see Norcia
\emph{et al.}, 2015, for an overview); however our focus here is on the
contrast response.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well (McKeefry \emph{et al.}, 1996; Di
Russo \emph{et al.}, 2001\emph{a}) (see also (Baseler \& Sutter, 1997)).

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}; note that the code used to produce all
figures in this review is available in python and R at:
https://github.com/wadelab/contrastReviewPaper).

@@

@@

To understand these processes better, we will show how sinusoidal input
signals might be processed by the visual system to produce SSVEPs.
Figure~\ref{fig-linear} illustrates how sine waves of different
contrasts are processed in a linear system. The first panel shows the
input sine wave, which would be used to modulate stimulus amplitude over
time. Notice that there are five peaks in the waveform during the one
second sample, so the stimulation frequency is 5Hz (F1). The second
panel shows Fourier transform of the waveform, which contains a
substantial peak at this frequency. If we change the stimulus contrast
(i.e.~the amplitude of the waveform), the amplitude of the F1 component
increases linearly with contrast (right panel).

@@

Next we can consider the impact of a nonlinearity in processing on the
responses. One of the simplest nonlinearities is the function that
describes a cell's response to different levels of contrast. In
Figure~\ref{fig-transducer} this is modeled by a hyperbolic ratio
function resulting in a saturating non-linearity:

@@

This nonlinearity has clear effects on signal transduction. The
sinusoidal waveform is distorted by the nonlinearity, and appears
frequency doubled (left panel of Figure~\ref{fig-transducer}; note that
other nonlinearities such as rectification and squaring can have similar
effects). The distortion is reflected in the Fourier spectrum (middle
panel of Figure~\ref{fig-transducer}), which now includes responses at
integer multiples (known as harmonics) of the stimulation frequency (10,
20 and 30Hz), but no response at the fundamental (5Hz). Finally, the
contrast response function is now nonlinear (right panel of
Figure~\ref{fig-transducer}). Although the hyperbolic ratio function
(Equation~\ref{eq-nakarushton}) is monotonic, the CRF resulting from
measuring the amplitude of the second harmonic (2F) component contains a
slight roll-off at high input contrasts. This results from the
distortion of the input sine waves at high contrast due to a combination
of the full-wave rectification and saturating non-linearity. Power at
other harmonics increases, and the total power generated by the input is
monotonic. This roll-off is often seen in experimental data and has been
referred to as `supersaturation' (Tyler \& Apkarian, 1985; Peirce,
2007).

@@

\includegraphics{Figures/transducer.pdf}

@@

\caption{\label{fig-transducer}Illustration of how a sinusoidal input
signal is affected by a nonlinear transducer. The input wavefore (left
panel) is frequency doubled by the squaring operator, and now features
energy at the second harmonic (10Hz) and its multiple (middle panel).
The contrast response function at the second harmonic is now nonlinear,
and saturates at high contrasts (right panel).}

@@

To illustrate the effect of contrast gain control (Heeger, 1992), we
next include a second component (a `mask') that contributes to the gain
pool of the first (the `target'). The mask stimulus will suppress the
response at the target frequency, reducing its amplitude. The
suppression is reciprocal - activity at the mask frequency is also
reduced by the presence of the target, in a contrast-dependent manner
(Busse \emph{et al.}, 2009). At a single pair of (matched) contrast
levels, we see a complex pattern of intermodulation terms in the
transduced waveforms (left panel of Figure~\ref{fig-twoinputs}) and in
the Fourier spectrum (middle panel of Figure~\ref{fig-twoinputs}). The
two components interact generating nonlinear intermodulation terms at
sums and differences of the input frequencies (F1, F2). The nature of
the interaction --- and therefore the pattern of intermodulation terms
--- is determined by the computations happening as the contrast signal
moves from retina to cortex (Regan \& Regan, 1988).

@@

\includegraphics{Figures/twoinputs.pdf}

@@

where \(C_\mathrm{mask}\) reflects the contrast of the mask component at
a distinct frequency from that of the target. The effect of this extra
term is to reduce the target response (see right panel of
Figure~\ref{fig-twoinputs}). For a linear contrast axis, the contrast
response function becomes shallower, whereas on a logarithmic contrast
axis it maintains its steepness and shifts to the right. Suppressive
effects of this kind have been obtained using SSVEP with a variety of
different types of mask, including orthogonal overlaid masks, surround
masks, and dichoptic masks (Burr \& Morrone, 1987; Ross \& Speed, 1991;
Candy \emph{et al.}, 2001; Busse \emph{et al.}, 2009; Cunningham
\emph{et al.}, 2017; Salelkar \& Ray, 2020; for a meta-analysis see
Baker \emph{et al.}, 2021).

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the LGN (Levitt
\emph{et al.}, 2001) and cortex (Hubel \& Wiesel, 1959; Movshon, 1975),
will introduce second harmonic components even when the stimulus itself
is modulated in an on-off fashion. Analogously, in the spatial domain,
so-called `simple cells' are sensitive to the polarity of a spatial
contrast modulation while `complex cells' respond to the presence of
patterned spatial contrast irrespective of its spatial position (Hubel
\& Wiesel, 1962). The SSVEP response to a contrast-reversing sine-wave
grating therefore contains information about nonlinear computations
performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging,
careful parametric variation of the input stimuli does allow us to fit
models of early visual gain control using SSVEP data. For example, Tsai
\emph{et al.} (2012) demonstrated that a gain control model gave a good
account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. Our more
recent work on signal combination across eyes and space similarly
demonstrated close correspondence between the predictions of a
computational model and empirical data in humans (Baker \& Wade, 2017).
More detailed modelling of intracortical recordings (Groen \emph{et
al.}, 2022) has revealed details of the timecourse of gain control
effects, specifically that normalization is delayed slightly relative to
the initial visual response.

Similar changes to the contrast response function might also be obtained
using adaptation paradigms, in which the visual system is exposed to
high contrast stimuli for long durations. Psychophysically, adaptation
increases detection thresholds, but has little effect on contrast
discrimination performance (Ross \emph{et al.}, 1993), much like pattern
masks (Foley, 1994). Although SSVEP adaptation effects show strong
tuning for orientation (Campbell \& Maffei, 1970; Vergeer \emph{et al.},
2018) and spatial frequency (Mecacci \& Spinelli, 1976), there appear to
be no studies measuring the full SSVEP contrast response function before
and after an extended period of adaptation.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels, and extrapolating back along the
function (either contrast vs response measured at a constant spatial
frequency or spatial frequency vs response at a constant contrast level)
to estimate its intercept with the x-axis (see
Figure~\ref{fig-sweepvep}). This contrast level was shown to correspond
approximately with psychophysically measured detection thresholds
(Norcia \emph{et al.}, 1986).

@@

\includegraphics{Figures/sweepplot.pdf}

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrast levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial (Tyler \emph{et al.}, 1979). To avoid
hysteresis effects, the sweep is sometimes conducted both up and down in
the same experiment (Tyler \emph{et al.}, 1979; Norcia \& Tyler,
1985\emph{a}; Norcia \emph{et al.}, 1990). The sweep VEP (really, a
sweep `SSVEP') technique is now commonly-used to obtain a rapid and
objective measurement of visual acuity. In particular, because of its
relative speed and simplicity, this technique has now become a standard
for conducting tests of visual acuity in very young subjects or where
behavioural tests are not appropriate (Ridder, 2004; Bach \emph{et al.},
2008; Hoffmann \emph{et al.}, 2017; Bach \& Farmer, 2020).

This approach has revealed much about the development of visual
abilities in infants (Harris \emph{et al.}, 1976; Atkinson \emph{et
al.}, 1979; Braddick \emph{et al.}, 1986). In general, SSVEP
measurements of infant vision have revealed that contrast sensitivity
for both achromatic and chromatic contrast as well as stereoscopic depth
perception develops earlier than had been supposed previously based on
behavioural readouts (Dobson \emph{et al.}, 1978; Norcia \& Tyler,
1985\emph{b}) with both chromatic and achromatic contrast detection
reaching near-adult levels by around six months. Spatial acuity as
measured by SSVEP reaches adult levels more slowly, but near-adult
levels are recorded around one year (Norcia \& Tyler, 1985\emph{b};
Hamer \emph{et al.}, 1989) compared to around six to seven years with
behavioural measures (Atkinson \& Braddick, 1983; Ellemberg \emph{et
al.}, 1999). At least some of this difference is likely due to the
relative objectivity and high SNR of the SSVEP technique compared to
other methods such as preferential looking, which require careful
measurement of the infant's gaze direction. However it should be noted
that other groups have reported electrophysiological correlates of
visual acuity that more closely match the behavioural measures (De
Vries-Khoe \& Spekreijse, 1982).

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)},
  pdfauthor={Alex R. Wade; Daniel H. Baker},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Measuring contrast processing in the visual system using the
steady state visually evoked potential (SSVEP)}
\author{Alex R. Wade \and Daniel H. Baker}
\date{}

@@

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for over a century. One of the most productive
experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high signal to noise ratio (SNR), and they allow us to
analyze interactions between stimulus components using a convenient
mathematical framework. Here we describe how SSVEPs have been used to
study visual contrast over the past 70 years (Dawson, 1954). Because our
thinking about SSVEPs is well-described by simple mathematical models,
we embed code that illustrates key steps in the modelling and analysis.
This paper can therefore be used both as a review of the use of SSVEP in
measuring human contrast processing, and as an interactive learning aid.

Keywords:

\begin{itemize}
\tightlist
\item
  EEG
\item
  VEP
\item
  SSVEP
\item
  Vision
\item
  Contrast
\end{itemize}

\newpage{}

@@

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as `contrast', sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity (Schade, 1956; Campbell \&
Green, 1965), defined as the inverse of the lowest contrast that can be
reliably detected. But neural responses can also be measured more
directly using techniques such as functional magnetic resonance imaging
(fMRI), magnetoencephalography (MEG), and electroencephalography (EEG).
Here we will describe how an EEG method known as the steady state
visually evoked potential (SSVEP) technique has contributed to our
understanding of human contrast processing in health, disease and
throughout development.

The SSVEP is a continuous electrical response evoked in the brain by
visual stimuli flickering at a constant frequency (Regan, 1966\emph{a}).
For contrast-defined stimuli, such as sine-wave gratings, it is
strongest at the occipital pole, adjacent to the early visual areas that
generate the signal, although careful analysis of individual
visually-evoked potentials (VEPs) reveals multiple generators throughout
visual cortex (Di Russo \emph{et al.}, 2005, 2007). The flickering
stimulus entrains neural population responses at multiples of the
stimulus frequency, so continuous EEG data are typically analysed by
taking the Fourier transform, and estimating the amplitude at these
frequencies.

Two common stimulus variants involve sinusoidal on-off (or
`appearance/disappearance') flicker, where the stimulus alternates
between a blank background and the peak contrast, and sinusoidal
counterphase flicker, where the stimulus alternates in phase (i.e the
black regions become white and the white regions become black). On-off
flicker can excite independent populations of on- and off-cells in the
visual system once per cycle. For spatial patterns, the contributions of
individual cortical cells to these excitations are thought to sum at the
scalp and generate an average of their onset responses on each cycle.
For very low spatial frequencies (including zero), resposes at the scalp
will be dominated by a single on- or off- cell type on each half cycle.

If the amplitudes and waveforms of responses to the stimulus appearance
and disappearnce are not perfectly balanced, a response at the
fundamental flicker frequency, known as 1F, and its integer harmonics
can be produced. For stimuli of any spatial frequency (including zero),
an imbalance can be caused by differences in either the onset- and
offset- responses of either on- or off-cells. For example, if only
on-sensitive neurons were present (firing to the onset, but not the
offset of local increases in luminance) the response to the on/off (or
`appearance/disappearance') presentation of a grating of any spatial
frequency below the resolution limit would consist entirely of odd
harmonics.

By comparison, counterphase flickering patterns generate two essentially
identical transients per cycle and therefore do not produce a response
at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves
are spectrally broad-band, square wave flicker (either on/off or
contrast reversing) tends to produce additional spectral harmonics
compared to sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). Because they can arise from different neuronal computations,
different populations and even different stages of the visual system,
different harmonics can exhibit different input-output functions. For
example, 1F and 2F responses can have different thresholds (Bobak
\emph{et al.}, 1984) and scalp topographies (Regan, 1973\emph{a}) while
Kaestner \emph{et al.} (2024) show both different response slopes and
thresholds for 1F and 2F components generated by dynamic noise.

SSVEP signals can also be elicited by periodic changes of stimulus
properties other than achromatic and chromatic contrast, such as motion,
stereo depth, and facial identity or expression (see Norcia \emph{et
al.}, 2015, for an overview); however our focus here is on the contrast
response.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well (Regan, 1966\emph{b}, 1973\emph{a},
1975; McKeefry \emph{et al.}, 1996; Di Russo \emph{et al.},
2001\emph{a}) (see also Baseler \& Sutter, 1997).

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Ohzawa \emph{et al.}, 1982, 1985; Bobak
\emph{et al.}, 1988; Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}; note that the code used to produce all
figures in this review is available in python and R at:
https://github.com/wadelab/contrastReviewPaperVNS).

@@

Although the SSVEP does seem to reflect the activity of relatively
well-tuned neuronal populations, changing some aspect of the stimulus
may change the nature of those populations. For example, Campbell \&
Maffei (1970) noted that densely-sampled measurements of contrast
responses can reveal the presence of two qualitatively different types
of neurons that exhibit different log-linear contrast functions
(Nakayama \& Mackeben, 1982; see also Tyler \& Apkarian, 1985; Norcia
\emph{et al.}, 1989; Souza \emph{et al.}, 2007). Modeling such
two-limbed contrast response functions using a single sigmoidal function
is therefore an approximation.

@@

An additional complexity is introduced by the fact that the EEG is an
average population response and that the visual system contains many
different types of neurons. As a stimulus changes in (for example)
contrast, it may selectively activate qualitatively different neuronal
populations.

To understand these processes better, we will show how sinusoidal input
signals might be processed by the visual system to produce SSVEPs.
Figure~\ref{fig-linear} illustrates how sine waves of different
contrasts are processed in a linear system. The first panel shows the
input sine wave, which would be used to modulate stimulus amplitude over
time. Notice that there are five peaks in the waveform during the one
second sample, so the stimulation frequency is 5Hz (F1). The second
panel shows the Fourier transform of the waveform, which contains a
substantial peak at this frequency. If we change the stimulus contrast
(i.e.~the amplitude of the waveform), the amplitude of the F1 component
increases linearly with contrast (right panel).

@@

Next we can consider the impact of two types of nonlinearity in
processing on the responses.

First, we consider the fact that neurons in the early visual system code
either positive or negative contrast. The response of each different
type of neuron is an increase in firing rate to the increasing levels of
the preferred contrast polarity, and these neurons typically exhibit
zero (or baseline) response to the non-preferred polarity. We can model
the combined responses of these two cell populations to a time varying
contrast signal by full-wave-rectification. The effect of this
computation is, effectively, to double the input frequency in the
population measurement and to introduce additional higher (even)
harmonics due to the discontinuity at the contrast reversals. The effect
of full-wave rectification on the SSVEP signal is shown in
Figure~\ref{fig-fullwaverectification}

\includegraphics{Figures/3_fullwaverectification.pdf}

\caption{\label{fig-fullwaverectification}Illustration of a full-wave
rectified sinusoid (left), its Fourier spectrum (middle), and how the
amplitude of the first harmonic component increases with contrast
(right).}

The contrast response function (CRF) is still linear for this model -
rectification has doubled the frequency (and introduced higher
harmonics) but the response amplitude measured by the second harmonic is
still directly proportional to the input amplitude.

Neurons in the visual system typically do not exhibit this perfect
response linearity. Instead, neuronal response are well-modeled by some
form of saturating non-linearity: as the contrast increases, the
response of the neuron increases but the response increase per unit
contrast is reduced at the high end of the contrast range. This
saturating nonlinearity is often modeled as a hyperbolic ratio function
of the form:

@@

This nonlinearity has clear effects on signal transduction. In
Figure~\ref{fig-transducer} the full-wave rectified waveform (which is
combining the response of both `on' and `off' neurons) passes through
the saturating nonlinearity (left panel of Figure~\ref{fig-transducer}).
The frequency doubling seen in Figure~\ref{fig-fullwaverectification} is
still present but the the contrast response function is now nonlinear
(right panel of Figure~\ref{fig-transducer}).

Interestingly, although the hyperbolic ratio function
(Equation~\ref{eq-nakarushton}) is monotonic, the CRF resulting from
measuring the amplitude of the second harmonic (2F) component contains a
slight roll-off at high input contrasts. This results from the
distortion of the input sine waves at high contrast due to a combination
of the full-wave rectification and saturating non-linearity. Power at
other harmonics increases, and the total power generated by the input is
monotonic. This roll-off is often seen in experimental data and has been
referred to as `supersaturation' (Tyler \& Apkarian, 1985; Peirce,
2007).

@@

\includegraphics{Figures/4_transducer.pdf}

@@

\caption{\label{fig-transducer}Illustration of how a sinusoidal input
signal is affected by a nonlinear transducer. The input waveform (left
panel) is frequency doubled by the full-wave rectification (see
Figure~\ref{fig-fullwaverectification}), and now has energy at the
second harmonic (10Hz) and higher even harmonics (middle panel). Passing
the signal through a second nonlinearity (the hyperbolic ratio function)
causes the contrast response function at the second harmonic (2F) to
become nonlinear and it and saturates (or even `supersaturates') at high
contrasts (right panel).}

@@

To illustrate the effect of contrast gain control (Ohzawa \emph{et al.},
1982; Heeger, 1992), we next include a second component (a `mask') that
contributes to the gain pool of the first (the `target'). The mask
stimulus will suppress the response at the target frequency, reducing
its amplitude. The suppression is reciprocal - activity at the mask
frequency is also reduced by the presence of the target, in a
contrast-dependent manner (Regan \& Regan, 1988; Brown \emph{et al.},
1999; Candy \emph{et al.}, 2001; Busse \emph{et al.}, 2009). At a single
pair of (matched) contrast levels, we see a complex pattern of
intermodulation terms in the transduced waveforms (left panel of
Figure~\ref{fig-twoinputs}) and in the Fourier spectrum (middle panel of
Figure~\ref{fig-twoinputs}). The two components interact generating
nonlinear intermodulation terms at sums and differences of the input
frequencies (F1, F2). The nature of the interaction --- and therefore
the pattern of intermodulation terms --- is determined by the
computations happening as the contrast signal moves from retina to
cortex (Regan \& Regan, 1988).

@@

\includegraphics{Figures/5_twoinputs.pdf}

@@

where \(C_\mathrm{mask}\) reflects the contrast of the mask component at
a frequency distinct from that of the target. The effect of this extra
term is to reduce the target response (see right panel of
Figure~\ref{fig-twoinputs}). For a linear contrast axis, the contrast
response function becomes shallower, whereas on a logarithmic contrast
axis it maintains its steepness and shifts to the right. Suppressive
effects of this kind have been obtained using SSVEP with a variety of
different types of mask, including orthogonal overlaid masks, surround
masks, and dichoptic masks (Burr \& Morrone, 1987; Ross \& Speed, 1991;
Candy \emph{et al.}, 2001; Busse \emph{et al.}, 2009; Cunningham
\emph{et al.}, 2017; Salelkar \& Ray, 2020; for a meta-analysis see
Baker \emph{et al.}, 2021).

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the lateral
geniculate nucleus (LGN) (Levitt \emph{et al.}, 2001) and cortex (Hubel
\& Wiesel, 1959; Movshon, 1975), will introduce second harmonic
components even when the stimulus itself is modulated in an on-off
fashion (McKeefry \emph{et al.}, 1996). Analogously, in the spatial
domain, so-called `simple cells' are sensitive to the polarity of a
spatial contrast modulation while `complex cells' respond to the
presence of patterned spatial contrast irrespective of its spatial
position (Hubel \& Wiesel, 1962). Both the amplitude and the phase of
the SSVEP response to a single contrast-reversing sine-wave grating
therefore contain information about nonlinear computations performed
across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along the entire processing pathway is challenging,
and a full system identification may require the use of multiple
different input frequencies (Boyd \emph{et al.}, 1983), careful
parametric variation of the input stimuli does allow us to fit models of
early visual visual processing (e.g. Zemon \& Gordon, 2006) and, by
incorporating mathematical models of gain control mechanisms with either
instantaneous, or temporally-integrated gain control, it is possible to
model gain control using SSVEP data. For example, Tsai \emph{et al.}
(2012) demonstrated that a relatively simple gain control model gave a
good account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. More recent
work has used this approach to study binocular integration in both
normal subjects and amblyopes (Hou \emph{et al.}, 2020; Hou 侯川
\emph{et al.}, 2021). Our more recent work on signal combination across
eyes and space similarly demonstrated close correspondence between the
predictions of a computational model and empirical data in humans (Baker
\& Wade, 2017). More detailed modelling of intracortical recordings
(Groen \emph{et al.}, 2022) has revealed details of the timecourse of
gain control effects, specifically that normalization is delayed
slightly relative to the initial visual response.

Similar changes to the contrast response function might also be obtained
using adaptation paradigms, in which the visual system is exposed to
high contrast stimuli for long durations. Psychophysically, adaptation
increases detection thresholds, but has little effect on contrast
discrimination performance (Ross \emph{et al.}, 1993), much like pattern
masks (Foley, 1994). Although SSVEP adaptation effects show strong
tuning for orientation (Campbell \& Maffei, 1970; Vergeer \emph{et al.},
2018) and spatial frequency (Mecacci \& Spinelli, 1976), there appear to
be very few measurements of the full SSVEP contrast response function
before and after an extended period of adaptation. One exception is a
study by Bach \emph{et al.} (1988) that examined the effects of
adaptation on SSVEP responses, and found a paradoxical increase in the
1F SSVEP amplitude following prolonged exposure to high-contrast stimuli
across much of the probe contrast range. They explain this as being due
to the presence of a nonlinear transducer function that saturates at
high probe contrasts, driving the response to higher harmonic
frequencies. Adaptation reduces the apparent contrast of the input,
leading to less distortion, a more sinusoidal output and a concentration
of the frequency-tagged response at lower harmonics. This explanation is
conceptually related to the `supersaturation' effect that we observed in
Figure 4.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels (in the linear part of the
log-contrast response function prior to saturation), and extrapolating
back along the function (either contrast vs response measured at a
constant spatial frequency or spatial frequency vs response at a
constant contrast level) to estimate its intercept with the x-axis
(Campbell \& Maffei, 1970) (see Figure~\ref{fig-sweepvep}). This
contrast level has been shown to correspond approximately with
psychophysically measured detection thresholds (Campbell \& Kulikowski,
1972; Norcia \emph{et al.}, 1986). One factor affecting the accuracy of
this technique is the presence of noise in the EEG signal which will
increase the apparent amplitude of the SSVEP responses across the
measurement range. The effect of this noise is particulary apparent for
recordings with low SNR or where the slope of the CRF is shallow (and
therefore where small changes in the vertical offset of the fit result
in large changes in the X-axis intercept). Accurate estimates of
thresholds therefore require the experimenter to model this noise and
compensate for its effects, as detailed in Norcia \emph{et al.} (1989).

@@

\includegraphics{Figures/6_sweepplot.pdf}

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrast levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial (Regan, 1973\emph{b}; Tyler \emph{et al.},
1979). To avoid hysteresis or short-term adaptation effects, the sweep
is sometimes conducted both up and down in the same experiment (Tyler
\emph{et al.}, 1979; Norcia \& Tyler, 1985\emph{a}; Norcia \emph{et
al.}, 1990). The sweep VEP (really, a `sweep SSVEP') technique is now
commonly-used to obtain a rapid and objective measurement of visual
acuity. In particular, because of its relative speed and simplicity,
this technique has now become a standard for conducting tests of visual
acuity in very young subjects or where behavioural tests are not
appropriate (Ridder, 2004; Bach \emph{et al.}, 2008; Hoffmann \emph{et
al.}, 2017; Bach \& Farmer, 2020).

The approach of estimating visual thresholds by extrapolating SSVEP
responses at higher stimulus levels has revealed much about the
development of visual abilities in infants (Harris \emph{et al.}, 1976;
Atkinson \emph{et al.}, 1979; Braddick \emph{et al.}, 1986). In general,
SSVEP measurements of infant vision have revealed that contrast
sensitivity and acuity for both achromatic and chromatic stimuli
develops earlier than had been supposed previously based on behavioural
readouts (Dobson \emph{et al.}, 1978; Norcia \& Tyler, 1985\emph{b}).
For example, using SSVEPs, Norcia \& Tyler (1985\emph{b}) found acuity
thresholds in one month old infants that were roughly half those
estimated previously from behavioural experiments (reviewed in Brown,
1990). The ability to make very sensitive measurements of early visual
function in infants also allows researchers to monitor the trajectory of
visual development and identify different temporal epochs of
developmental change (Norcia \emph{et al.}, 1990).

SSVEP measures of thresholds for detecting purely chromatic stimuli find
responses to isoluminant stimuli by around 5-6 weeks postnatally, with
contrast detection reaching near-adult levels by around six months
(Morrone \emph{et al.}, 1993). Spatial acuity as measured by SSVEP
reaches adult levels more slowly, but near-adult levels are recorded
around one year (Norcia \& Tyler, 1985\emph{b}; Hamer \emph{et al.},
1989) compared to around six to seven years with behavioural measures
(Atkinson \& Braddick, 1983; Ellemberg \emph{et al.}, 1999).

At least some of this difference is likely due to the relative
objectivity and high SNR of the SSVEP technique compared to other
methods such as preferential looking, which requires careful measurement
of the infant's gaze direction. In slightly older chilren, other groups
have reported electrophysiological correlates of visual acuity that more
closely match the behavioural measures (De Vries-Khoe \& Spekreijse,
1982).
